---
title: "GSEA analysis"
author: "Nick Diercksen"
date: "Last compiled on `r format(Sys.time())`"
output:
  html_notebook:
    toc: yes
    toc_float: yes
  html_document:
    toc: yes
    df_print: paged
editor_options:
  chunk_output_type: inline
params:
  reevaluate:
    label: Should all calculations and database lookups be done again or just e.g.
      graphics regenerated
    value: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
# library(ggplot2)
library(rWikiPathways)
library(xml2)
library(kableExtra)
```

```{r load_rdata, include=FALSE}
get_xml_text <- function(nodes, xpath) {
    xml_find_all( nodes, xpath ) %>%
    xml_text() %>% 
    ifelse(length(.) == 0, "", .)
}
extract_metaInf_from_gpml <- function(gpml_file) {
  doc <- read_xml(gpml_file)
  # doc1 <- read_xml(gpml_files[119]) # Microglia pathogen phagocytosis pathway
  
  pathway_name <- xml_attr(doc, "Name")
  
  # "Pathway Ontology" and "Cell Type" information:
  if ("bp" %in% names(xml_ns(doc))) {
    # (throws error if namespace 'bp' does not exist)
    ontology_nodes <-
      xml_find_all(doc, "//d1:Biopax")
    pathway_ontology <-
      get_xml_text(
        ontology_nodes,
        "//bp:openControlledVocabulary[bp:Ontology='Pathway Ontology']/bp:TERM"
      )
    
    cell_type <-
      get_xml_text(
        ontology_nodes,
        "//bp:openControlledVocabulary[bp:Ontology='Cell Type']/bp:TERM"
      )
  } else {
    pathway_ontology <- ""
    cell_type <- ""
  }
  
  source_nodes <- xml_find_all(doc, "./*[@Source]")
  description <-
    get_xml_text(source_nodes, "//*[@Source='WikiPathways-description']")
  
  category <-
    get_xml_text(source_nodes, "//*[@Source='WikiPathways-category']")
  
  df <- data.frame(
    "Pathway" = pathway_name,
    "Category" = category,
    "Pathway Ontology" = pathway_ontology,
    "Cell Type" = cell_type,
    "Description" = description
  )
  
  return(df)
}
getWikiPathways <- function() {
  
  pathways.dir <- "./data/pathways/"
  wikiPathways.dir <- "./data/pathways/wikiPathways"
  wikiPathways.file <- "./data/pathways/wikiPathways.rds"
  if (file.exists(wikiPathways.file) && !params$reevaluate) {
    return(readRDS(wikiPathways.file))
  }
  
  if (!file.exists(wikiPathways.dir)) {
    dir.create(wikiPathways.dir)
  }
  
  # download gpml files from wikiPathways
  
  tmp_dir <- file.path(tempdir(), "wikipathways")
  dir.create(tmp_dir)
  wikiPathway.zip <-
    rWikiPathways::downloadPathwayArchive(organism = "Mus musculus",
                                          format = 'gpml',
                                          destpath = tmp_dir)
  
  gpml_files <-
    unzip(file.path(tmp_dir, wikiPathway.zip), exdir = tmp_dir)
  
  df_list <- lapply(gpml_files, function(f) {
    extract_metaInf_from_gpml(f)
  })
  
  # chain merge into final dataframe
  final_df <-
    Reduce(function(x, y)
      merge(x, y, all = T), df_list)
  
  saveRDS(final_df, wikiPathways.file)
  
  unlink(tmp_dir, recursive = TRUE)
  return(final_df)
}

wikiPathways <- getWikiPathways()
fgseaRes.combined <-
  readRDS("./data/Robjects/04_fgseaRes.combined.rds") %>%
  mutate(NES = abs(NES.ga) + abs(NES.sol))
```

# interpretting significant pathways
Goal here is to get an understanding of the significant pathways from the fgsea
results.

Collecting metadata for the pathways is one of the first steps.
[wikiPathways](https://www.wikipathways.org/index.php/Download_Pathways) is one
of the external links from the [msigdb](http://www.gsea-msigdb.org/gsea/msigdb/mouse/collections.jsp)
entries. It only provides information to `r nrow(wikiPathways)` pathways and even these are not


## wikiPathways (metadata)
```{r}
kableExtra::kbl(wikiPathways) %>%
  kableExtra::kable_paper() %>% 
  kableExtra::scroll_box(width = "100%", height = "1000px")
```

# currentTODOs

[ ] find out biological meaning of significant pathways\
[ ] merge `wikiPathways` with the `fgseaRes.combined` to get a better overview
